COFFEE: an Optimizing Compiler\\for Finite Element Local Assembly

### 2014-04-11 paper/bare_conf.tex
Incorporate Ana's suggestions

We focus on local assembly (``assembly'', in the following), a fundamental step of a finite element method that covers an important fraction of the overall computation run-time, often in the range 30$\%$-60$\%$. During the assembly phase, the solution of the PDE is approximated by executing a suitable kernel over all elements in the discretized domain. A kernel's working set is usually small enough to fit the L1 cache; it might need L2 cache when high-order methods are employed to improve the accuracy of the solution. However, we do not consider the latter case. An assembly kernel is characterized by the presence of an affine, often non-perfect loop nest, where individual loops are rather small: their trip count rarely exceeds 30, and may be as low as 3 for lower order methods. In the innermost loop, a problem-specific expression evaluates a two dimensional array, representing the result of local assembly in an element of the discretized domain. With such a kernel structure, we focus on aspects like minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorization.

Achieving high-performance is non-trivial. The complexity of the mathematical expressions, often characterized by a large number of operations on constants and small matrices, makes it hard to determine a single or specific sequence of transformations that is successfully applicable to all problems. The variation in loop trip counts and their typically small value, which prevents the adoption of standard techniques for cache locality, further exacerbates the issue. A compiler-based approach is, therefore, the only reasonable option to obtain close-to-peak performance in a wide range of different local assembly kernels. Optimizations like generalized loop-invariant code motion, vector-register tiling, and expression splitting, as well as their composition, are essential, but they are not supported by state-of-the-art polyhedral and vendor compilers. BLAS routines could be theoretically employed, although a fairly complicated control- and data-flow analysis would be required to automate identification and extraction of matrix-matrix multiplies. In addition, as detailed in Section~\ref{sec:perf-eval-blas}, the small dimension of the involved matrices and the potential loss in data locality can limit the performance gain, if any.

Due to the constraints of available compilers and linear-algebra specialized libraries, we have automated a set of generic and model-driven code transformations in COFFEE\footnote{COFFEE stands for COmpiler For FinitE Element local assembly.}, a compiler for optimizing local assembly kernels. COFFEE is integrated with Firedrake~\cite{firedrake-code}, a system for solving PDEs through the finite element method based on the PyOP2 abstraction~\cite{pyop2isc}. It supports any problems expressible with this framework. This allows us to evaluate our code transformations in a range of real-world problems, varying key parameters that impact both solution accuracy and kernel cost, namely the polynomial order of the method (from $p=1$ to $p=4$) and the geometry of elements in the discretized domain (2D triangle, 3D tetrahedron, 3D prism).

Early experiments showed that Firedrake-generated code for non-trivial assembly kernels was sub-optimal. Our cost-model-driven sequence of source-to-source code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 1.5$\times$ over ``softly-optimized'' code (i.e. code where only basic transformations are performed, such as generalized loop-invariant code motion, padding, and data alignment), and up to 4.44$\times$ over original kernels. The contributions of this paper are threefold
\begin{itemize}
\item An optimisation strategy for finite element local assembly. Our approach exploits domain knowledge and goes beyond the limits of both vendor and research compilers.
\item Design and implementation of a compiler that automates the proposed code transformations for any problems expressible in Firedrake.
\item Systematic evaluation using a suite of examples of real-world importance, and evidence of significant performance improvements on two Intel architectures, a Sandy Bridge CPU and the Xeon Phi.
\end{itemize}

@@

Given a mathematical description of the input problem, expressed through the domain-specific Unified Form Language~\cite{ufl}, Firedrake generates C-code kernels implementing assembly using a numerical integration algorithm. It then triggers compilation of such kernels using an available vendor compiler, and eventually manages parallel execution over the mesh. The subject of this paper is to enhance this execution model by adding an optimization stage prior to the generation of C code. The code transformations described next are also generalizable to non-Firedrake assembly kernels, provided that numerical integration is used.

The general structure of a Firedrake-generated kernel is shown in Figure~\ref{code:general-structure}. Input are a pointer to a zero-initialized two dimensional array used to store the element matrix, a pointer to the element's coordinates in the discretized domain, and pointers to data of coefficient fields, for instance indicating value of velocity or pressure in the element. The output is the evaluated element matrix. A local assembly kernel can be logically split into three parts: 1) calculation of the Jacobian matrix, its determinant and its inverse; 2) definition of some constant two dimensional arrays, called ``basis functions'' (and their derivatives) in finite element terminology, initialized to problem-specific double-precision floating point values; 3) evaluation of the element matrix in an affine loop nest. All elements in the discretized domain share the same basis function arrays, so they are declared as global read-only arrays (i.e. using \texttt{static const} in C language). Table~\ref{table:map-name-letters} shows the variable names we will use in the upcoming code snippets to refer to the various kernel's objects.

\begin{table}[h]
\begin{center}
\begin{tabulary}{1.0\columnwidth}{|C|C|C|}
\hline
Object name & Type & Variable name(s) \\\hline\hline
Determinant of Jacobian matrix & double & det  \\ \hline
Inverse of Jacobian matrix & double[] & K \\ \hline
Coordinates & double** & coords\\ \hline
Fields & double** & w \\ \hline
Integration points & double[] & W \\ \hline
Basis functions (and derivatives) & double[][] & X, Y, X1, ... \\ \hline
Element matrix & double[][] & A\\ \hline
\end{tabulary}
\end{center}
\caption{Type and variable names used in the various code snippets to identify local assembly objects.}
\label{table:map-name-letters}
\end{table}

The actual complexity of a local assembly kernel depends on the finite element problem being solved. In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products involving basis functions. An example is provided in Figure~\ref{code:helmholtz}, which shows the assembly kernel for a Helmholtz problem using Lagrange basis functions on 2D elements with polynomial order $p=1$. In other scenarios, for instance when solving a non-linear problem, like Burgers in Figure~\ref{code:burgers}, the number of arrays involved in the computation of the element matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression. Also, constants evaluated in outer loops (called $F$ in the code), acting as scaling factors of arrays, may be required; trip counts can be larger (proportionally to the order of the method); arrays may be block-sparse. Note that in addition to a larger number of operations to compute the element matrix, the Burgers case shows a register pressure higher than that in Helmholtz. Despite assembly kernels being problem-dependent, meaning that the space of codes that Firedrake can generate is infinite, it is still possible to identify common domain-specific traits that can be exploited for effective code transformations and SIMD vectorization.

Some peculiarities of assembly kernels are: 1) memory accesses along the three loop dimensions are always stride-1; 2) the $j$ and $k$ loops are interchangeable, whereas permutation of $i$ might be subjected to pre-computation of values (e.g. the $F$ values in Burgers) and introduction of temporary arrays; 3) the $j$ and $k$ loops iterate over the same iteration space; 4) most of the sub-expressions on the right hand side of the element matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$). In Section~\ref{sec:code-transf} we show how to exploit these observations to define a set of systematic, composable optimizations.

\section{Code Transformations}
\label{sec:code-transf}
The code transformations presented in this section are applicable to all finite element problems that can be formulated in Firedrake. One peculiar characteristic is that they all aim at improving the run-time of the \texttt{ijk} loop nest, where the numerical integration takes place. In rare, yet important cases, however, this might not be sufficient to achieve notable performance improvements. Generalized loop-invariant code motion, which is described in Section~\ref{sec:code-transf-licm}, is a fundamental optimization that allows pre-computation of invariant sub-expressions. It is worth noting there are circumstances in which the amount of lifted terms, either at the level of an outer loop or even completely outside of the loop nest, is so big (thousands of operations, hundreds of temporaries) that their execution time becomes the dominant factor of the overall local assembly run-time. In these cases, two challenging optimizations are common sub-expression elimination and effective vectorization (superword level parallelism~\cite{SLP} would not be of help, because invariant sub-expressions lack, in general, structure). Given its inherent complexity, a comprehensive study of this problem is left as further work.

As already emphasized, the structure of mathematical expressions evaluating the element matrix and the variation in loop trip counts, although typically limited to the order of tens iterations, render the optimization process challenging. Relatively to machine peak, while some problems achieve good performance when applying a very specific set of transformations, there is also an ample space of cases for which a more elaborated optimization strategy is required. For example, the Burgers problem in Figure~\ref{code:burgers}, given the large number of arrays accessed, suffers from high register pressure, whereas the Helmholtz problem in Figure~\ref{code:helmholtz} does not; this intuitively suggests that the two problems require a different treatment, based on an in-depth analysis of both data and iteration spaces. Furthermore, domain-knowledge enables transformations that a general-purpose compiler could not apply, making the optimization space even larger. In this context, our goal is to understand the relationship between distinct code transformations, their impact on local assembly kernels, and to what extent their composability is effective in a class of problems and architectures.

% Padding and Data Alignment
\subsection{Prerequisites for Effective Auto-vectorization: Padding and Data Alignment}
Auto-vectorization of assembly code computing the element matrix can be less effective if data are not aligned and if the length of the innermost loop is smaller than the vector length $vl$. Data alignment is enforced in two steps. Initially, all arrays are allocated to addresses that are multiples of $vl$. Then, two dimensional arrays are padded by rounding the number of columns to the nearest multiple of $vl$. For instance, assume the original size of a basis function array is 3$\times$3 and that the underlying architecture possesses AVX, which implies $vl=4$ since a vector register is 256 bits long and our kernels use 64-bits double-precision floating-point values. In this example, a padded version of the array will have size 3$\times$4. The compiler is explicitly informed about data alignment using a suitable pragma. Padding of all two dimensional arrays involved in the evaluation of the element matrix also allows us to safely round the loop trip count to the nearest multiple of $vl$. This avoids the introduction of a remainder (scalar) loop from the compiler, which would render vectorization less efficient.

% Autovectorizable LICM -> show code snippet?
\subsection{Generalized Loop-invariant Code Motion}
\label{sec:code-transf-licm}
From inspection of the codes in Figures~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions that depend on two iteration variables only. Since symbols in most of these sub-expressions are read-only variables, there is ample space for loop-invariant code motion. Vendor compilers apply this technique, although not in the systematic way we need for our assembly kernels. We want to overcome two deficiencies that both $intel$ and $gnu$ compilers have. First, they can identify sub-expressions that are invariant with respect to the innermost loop only. This is an issue for sub-expressions depending on $i$-$k$, which are not automatically lifted in the loop order \texttt{ijk}. Second, the hoisted code is scalar, i.e. it is not subjected to auto-vectorization. We work around these limitations with source-level loop-invariant code motion. In particular, we pre-compute all values that an invariant sub-expression assumes along its fastest varying dimension. This is implemented by introducing a temporary array per invariant sub-expression and by adding a new loop to the nest. At the price of extra memory for storing temporaries, the gain is that lifted terms can be auto-vectorized, because part of an inner loop. Given the short trip counts of our loops, it is important to achieve auto-vectorization of hoisted terms in order to minimize the percentage of scalar instructions, which could be otherwise significant. It is also worth noting that, in some problems, invariant sub-expressions along $j$ are identical to those along $k$ (e.g. in Helmholtz). In these cases, we safely avoid redundant pre-computation since, as anticipated in Section~\ref{sec:background}, a property of our domain is that $j$ and $k$ loops share the same iteration space.

@@

% Register allocation and register locality
\subsection{Model-driven Vector-register Tiling}
One notable problem of assembly kernels concerns register allocation and register locality. The critical situation occurs when loop trip counts and accessed variables are such that the vector-registers pressure is high. Since the kernel's working set fits the L1 cache, it is remarkably important to optimize register management. Canonical optimizations, such as loop interchange, unroll, and unroll-and-jam, can be employed to deal with this problem. In COFFEE, these optimizations are supported either by means of explicit code transformations (interchange, unroll-and-jam) or indirectly by delegation to the compiler through standard pragmas (unroll). Tiling at the level of vector registers is an additional feature of COFFEE. Based on the observation that the evaluation of the element matrix can be reduced to a ``summation of outer products'' along the $j$ and $k$ dimensions, a model-driven vector-register tiling strategy can be implemented. If we consider the code snippet in Figure~\ref{code:helmholtz-licm} and we ignore the presence of the operation \texttt{det*W3[i]}, we can notice that the computation of the element matrix is abstractly expressible as
\begin{equation}
\label{outer-product}
A_{jk} = \sum_{\substack{
  x \in B' \subseteq B \\
  y \in B'' \subseteq B}}
x_j\cdot y_k ~~~~~~ j,k = 0,...,4
\end{equation}
where $B$ is the set of all basis functions (or temporary variables, e.g. \texttt{LI$\_$0}) accessed in the kernel, whereas $B'$ and $B''$ are generic problem-dependent subsets. Regardless of the specific input problem, and by abstracting from the presence of all variables independent of both $j$ and $k$, the element matrix computation is always reducible to this form. Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 entries ($j,k=0,...,4$) of the element matrix using just 2 vector registers, which represent a 4$\times$4 tile, assuming $\vert B' \vert = \vert B'' \vert = 1$. Values in a register are shuffled each time a product is performed. Standard compiler auto-vectorization ($gnu$ and $intel$), instead, executes 4 broadcast operations (i.e. ``splat'' of a value over all of the register locations) along the outer dimension to perform the calculation. Besides a larger number of cache accesses, it needs to keep between $f=1$ and $f=3$ extra registers to perform the same 16 evaluations when unroll-and-jam is used, with $f$ being the unroll-and-jam factor.

@@

\begin{figure}[t]
\centerline{\includegraphics[scale=0.7]{Pictures/vect-restore.pdf}}
\caption{Restoring the storage layout after \emph{op-vect}. The figure shows how 4$\times$4 elements in the top-left block of the element matrix $A$ can be moved to their correct positions. Each rotation, represented by a group of three same-colored arrows, is implemented by a single vector shuffle intrinsics.}
\label{fig:restore-layout}
\end{figure}

@@

% Splitting
\subsection{Expression Splitting}
\label{sec:expr-split}
In complex kernels, like Burgers' in Figure~\ref{code:burgers}, and on certain architectures, achieving effective register allocation can be challenging. If the number of variables independent of the innermost-loop dimension is close to or greater than the number of available CPU registers, it is likely to obtain poor register reuse. This usually happens when the number of basis function arrays, temporaries introduced by generalized loop-invariant code motion, and problem constants is large. For example, applying loop-invariant code motion to Burgers on a 3D mesh needs 33 temporaries for the \texttt{ijk} loop order, and compiler's hoisting of invariant loads out of the $k$ loop can be inefficient on architectures with a relatively low number of registers. One potential solution to this problem consists of suitably ``splitting'' the computation of the element matrix $A$ into multiple sub-expressions; an example, for the Helmholtz problem, is given in Figure~\ref{code:helmholtz-split}. Splitting an expression has, in general, several drawbacks. Firstly, it increases the number of accesses to $A$ proportionally to the ``split factor'', which is the number of sub-expressions produced. Secondly, depending on how the split is executed, it can lead to redundant computation (e.g. the product $det*W3[i]$ is performed times number of sub-expressions in the code of Figure~\ref{code:helmholtz-split}). Finally, it might affect register locality, although this is not the case of the Helmholtz example: for instance, the same array could be accessed in different sub-expressions, requiring a proportional number of loads be performed. Nevertheless, as shown in Section~\ref{sec:perf-results}, the performance gain from improved register reuse along inner dimensions can still be greater, especially if the split factor and the splitting itself use heuristics to minimize the aforementioned issues.

@@

Firedrake provides users with the Unified Form Language to write problems in a notation resembling mathematical equations. At run-time, the high-level specification is translated by the FEniCS Form Compiler~\cite{FFC-Compiler} into an abstract syntax tree (AST) representation of one or more finite element assembly kernels. ASTs are then passed to COFFEE, which is capable of applying the transformations described in Section~\ref{sec:code-transf}. The output of COFFEE, C code, is eventually provided to PyOP2~\cite{pyop2isc}, where just-in-time compilation and execution over the discretized domain take place. Because of the large number of (possibly parametric) transformations, COFFEE needs a mechanism to select the most suitable optimization strategy for a given problem. Auto-tuning might be used, although it would significantly increase the run-time overhead, since the generation of ASTs occurs at run-time as soon as problem-specific data are available. Our optimization strategy, based on heuristics and a simple cost model, is described in the following, along with an overview of the compiler.

@@

All loops are interchangeable provided that temporaries are introduced if the nest is not perfect. For the employed storage layout, the loop permutations \texttt{ijk} and \texttt{ikj} are likely to maximize performance. Conceptually, this is motivated by the fact that if the $i$ loop were in an inner position, then a significantly higher number of load instructions would be required for every iteration. Experiments showed that the performance loss is greater than the gain due to the possibility of accumulating increments in a register, rather than in memory, along the $i$ loop. The choice between \texttt{ijk} and \texttt{ikj} depends on the number of load instructions that can be hoisted out of the innermost dimension. Our compiler chooses the loop along which the number of invariant loads is smaller as outer so that more registers remain available to carry out the computation of the element matrix.

Loop unroll and unroll-and-jam of outer loops are fundamental to expose instruction-level parallelism, so tuning critical parameters, like the unroll factor, is of great importance. However, from our experience of inspecting assembly code and comparing with other hand-written implementations, recent versions of a vendor compiler like Intel's employ cost models capable of estimating close-to-optimal values for such parameters. This is particularly true for assembly kernels, where the loop nest is affine, bounds are usually very small and known at compile-time, and memory accesses are unit-stride. We therefore leave the backend compiler in charge of selecting the unroll factor. This choice also simplifies the COFFEE's cost model. The only situation in which we explicitly unroll-and-jam a loop is when \emph{op-vect} is used, since the transformed code seems to prevent the Intel compiler from applying this optimization, even if specific pragmas are added. Note that, regardless of the output of the cost model, the unroll-and-jam factor never exceeds the actual size of the outer loop (line 19 in Figure~\ref{algo:PyOP2Compiler}, assuming the default loop order \texttt{ijk}).

>>>>

We focus on improving the performance of local assembly (``assembly'', in the following), a fundamental step of a finite element method that covers an important fraction of the overall computation run-time, often in the range 30$\%$-60$\%$. During the assembly phase, the solution of the PDE is approximated by executing a suitable kernel over all elements in the discretized domain. A kernel's working set is usually small enough to fit the L1 cache; it might need L2 cache when high-order methods are employed to improve the accuracy of the solution. However, we do not consider the latter case. An assembly kernel is characterized by the presence of an affine, often non-perfect loop nest, where individual loops are rather small: their trip count rarely exceeds 30, and may be as low as 3 for lower order methods. In the innermost loop, a problem-specific expression evaluates a two dimensional array, representing the result of local assembly in an element of the discretized domain. With such a kernel structure, we focus on aspects like minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorization.

Achieving high-performance is non-trivial. The complexity of the mathematical expressions, often characterized by a large number of operations on constants and small matrices, makes it hard to determine a single or specific sequence of transformations that is successfully applicable to all problems. The variation in loop trip counts and their typically small value further exacerbates the issue. A compiler-based approach is, therefore, the only reasonable option to obtain close-to-peak performance in a wide range of different local assembly kernels. Optimizations like generalized loop-invariant code motion, vector-register tiling, and expression splitting, as well as their composition, are essential, but none are supported by state-of-the-art polyhedral and vendor compilers. BLAS routines could be theoretically employed, although a fairly complicated control- and data-flow analysis is required to automate identification and extraction of matrix-matrix multiplies. In addition, as detailed in Section~\ref{sec:perf-eval-blas}, the small dimension of the involved matrices and the potential loss in data locality can limit the performance gain, if any.

Due to the constraints of available compilers and linear-algebra specialized libraries, we have automated a set of generic and model-driven code transformations in COFFEE\footnote{COFFEE stands for COmpiler For FinitE Element local assembly.}, a compiler for optimizing local assembly kernels. COFFEE is integrated with Firedrake~\cite{firedrake-code}, a system for solving PDEs through the finite element method based on the PyOP2 abstraction~\cite{pyop2isc}. It supports any problems expressible with this framework. This allows us to evaluate our code transformations in a range of real-world problems, varying two key parameters that impact both solution accuracy and kernel cost: the polynomial order of the method (from $p=1$ to $p=4$) and the geometry of elements in the discretized domain (2D triangle, 3D tetrahedron, 3D prism).

Our experiments showed that the originally generated code for non-trivial assembly kernels was sub-optimal. Our cost-model-driven sequence of source-to-source code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 1.5$\times$ over ``softly-optimized'' code (i.e. code where only basic transformations are performed, such as generalized loop-invariant code motion, padding, and data alignment), and up to 4.4$\times$ over original kernels. The contributions of this paper are threefold
\begin{itemize}
\item An optimisation strategy for finite element local assembly. Our approach exploits domain knowledge and goes beyond the limits of both vendor and research compilers.
\item Design and implementation of a compiler that automates the proposed code transformations for any problems expressible in Firedrake.
\item Systematic analysis using a suite of examples of real-world importance, and evidence of significant performance improvements on two Intel architectures, a Sandy Bridge CPU and the Xeon Phi.
\end{itemize}

@@

Given a mathematical description of the input problem, expressed through the domain-specific Unified Form Language~\cite{ufl}, Firedrake generates C-code kernels implementing assembly using a numerical integration algorithm. It then triggers compilation of such kernels using a vendor compiler, and eventually manages parallel execution over the mesh. The subject of this paper is to enhance this execution model by adding an optimization stage prior to the generation of C code. The code transformations described next are also generalizable to non-Firedrake assembly kernels, provided that numerical integration is used.

The general structure of a Firedrake-generated kernel is shown in Figure~\ref{code:general-structure}. Input are a zero-initialized two dimensional array used to store the element matrix, the element's coordinates in the discretized domain, and coefficient fields, for instance indicating value of velocity or pressure in the element. The output is the evaluated element matrix. A local assembly kernel can be logically split into three parts: 1) calculation of the Jacobian matrix, its determinant and its inverse; 2) definition of some constant two dimensional arrays, called ``basis functions'' (and their derivatives) in finite element terminology, initialized to problem-specific double-precision floating point values; 3) evaluation of the element matrix in an affine loop nest. All elements in the discretized domain share the same basis function arrays, so they are declared as global read-only arrays (i.e. using \texttt{static const} in C). Table~\ref{table:map-name-letters} shows the variable names we will use in the upcoming code snippets to refer to the various kernel's objects.

\begin{table}[h]
\begin{center}
\begin{tabulary}{1.0\columnwidth}{|C|C|C|}
\hline
Object name & Type & Variable name(s) \\\hline\hline
Determinant of the Jacobian matrix & double & det  \\ \hline
Inverse of the Jacobian matrix & double[] & K \\ \hline
Coordinates & double** & coords\\ \hline
Fields & double** & w \\ \hline
Integration points & double[] & W \\ \hline
Basis functions (and derivatives) & double[][] & X, Y, X1, ... \\ \hline
Element matrix & double[][] & A\\ \hline
\end{tabulary}
\end{center}
\caption{Type and variable names used in the various code snippets to identify local assembly objects.}
\label{table:map-name-letters}
\end{table}

The actual complexity of a local assembly kernel depends on the finite element problem being solved. In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products involving basis functions. An example is provided in Figure~\ref{code:helmholtz}, which shows the assembly kernel for a Helmholtz problem using Lagrange basis functions on 2D elements with polynomial order $p=1$. In other scenarios, for instance when solving a non-linear problem, like Burgers in Figure~\ref{code:burgers}, the number of arrays involved in the computation of the element matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression. Also, constants evaluated in outer loops (called $F$ in the code), acting as scaling factors of arrays, may be required; trip counts can be larger (proportionally to the order of the method); arrays may be block-sparse. Note that in addition to a larger number of operations to compute the element matrix, the Burgers case shows a register pressure higher than that in Helmholtz.

Despite assembly kernels being problem-dependent, meaning that the space of codes that Firedrake can generate is infinite, it is still possible to identify common domain-specific traits that can be exploited for effective code transformations and SIMD vectorization. These include: 1) memory accesses along the three loop dimensions are always stride-1; 2) the $j$ and $k$ loops are interchangeable, whereas permutation of $i$ might be subjected to pre-computation of values (e.g. the $F$ values in Burgers) and introduction of temporary arrays; 3) the $j$ and $k$ loops iterate over the same iteration space; 4) most of the sub-expressions on the right hand side of the element matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$). In Section~\ref{sec:code-transf} we show how to exploit these observations to define a set of systematic, composable optimizations.

\section{Code Transformations}
\label{sec:code-transf}
The code transformations presented in this section are applicable to all finite element problems that can be formulated in Firedrake. One peculiar characteristic is that they all aim at improving the run-time of the \texttt{ijk} loop nest, where the numerical integration takes place. In rare cases (e.g. problems based on Hyperelasticity), however, such optimizations might not be sufficient to achieve notable performance improvements. Generalized loop-invariant code motion, which is described in Section~\ref{sec:code-transf-licm}, is a fundamental optimization that allows pre-computation of invariant sub-expressions. It is worth noting there are circumstances in which the amount of lifted terms, either at the level of an outer loop or even completely outside of the loop nest, is so big (thousands of operations, hundreds of temporaries) that their execution time becomes the dominant factor of the overall local assembly run-time. In these cases, two challenging optimizations are common sub-expression elimination and effective vectorization (superword level parallelism~\cite{SLP} would not be of help, because invariant sub-expressions lack, in general, structure). A comprehensive study of such cases is left as further work.

As already emphasized, the structure of mathematical expressions evaluating the element matrix and the variation in loop trip counts, although typically limited to the order of tens iterations, render the optimization process challenging. It is not always the same set of optimizations that bring performance closest to the machine peak. For example, the Burgers problem in Figure~\ref{code:burgers}, given the large number of arrays accessed, suffers from high register pressure, whereas the Helmholtz problem in Figure~\ref{code:helmholtz} does not; this intuitively suggests that the two problems require a different treatment, based on an in-depth analysis of both data and iteration spaces. Furthermore, domain-knowledge enables transformations that a general-purpose compiler could not apply, making the optimization space even larger. In this context, our goal is to understand the relationship between distinct code transformations, their impact on local assembly kernels, and to what extent their composability is effective in a class of problems and architectures.

% TODO: could add a sentence here to bridge the two paragraphs

% Padding and Data Alignment
\subsection{Prerequisites for Effective Auto-vectorization: Padding and Data Alignment}
Auto-vectorization of assembly code computing the element matrix can be less effective if data are not aligned and if the length of the innermost loop is smaller than the vector length $vl$. Data alignment is enforced in two steps. Initially, all arrays are allocated to addresses that are multiples of $vl$. Then, two dimensional arrays are padded by rounding the number of columns to the nearest multiple of $vl$. For instance, assume the original size of a basis function array is 3$\times$3 and $vl=4$ (e.g. AVX processor, with 256 bits long vector registers and 64-bits double-precision floating-point values). In this case, a padded version of the array will have size 3$\times$4. The compiler is explicitly informed about data alignment using a suitable pragma. Padding of all two dimensional arrays involved in the evaluation of the element matrix also allows us to safely round the loop trip count to the nearest multiple of $vl$. This avoids the introduction of a remainder (scalar) loop from the compiler, which would render vectorization less efficient.

% Autovectorizable LICM -> show code snippet?
\subsection{Generalized Loop-invariant Code Motion}
\label{sec:code-transf-licm}
From the inspection of the codes in Figures~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions that depend on two iteration variables only. Since symbols in most of these sub-expressions are read-only variables, there is ample space for loop-invariant code motion. Vendor compilers apply this technique, although not in the systematic way we need for our assembly kernels. We want to overcome two deficiencies that both $intel$ and $gnu$ compilers have. First, they can identify sub-expressions that are invariant with respect to the innermost loop only. This is an issue for sub-expressions depending on $i$-$k$, which are not automatically lifted in the loop order \texttt{ijk}. Second, the hoisted code is scalar, i.e. it is not subjected to auto-vectorization. We work around these limitations with source-level loop-invariant code motion. In particular, we pre-compute all values that an invariant sub-expression assumes along its fastest varying dimension. This is implemented by introducing a temporary array per invariant sub-expression and by adding a new loop to the nest. At the price of extra memory for storing temporaries, the gain is that lifted terms can be auto-vectorized, as part of an inner loop. Given the short trip counts of our loops, it is important to achieve auto-vectorization of hoisted terms in order to minimize the percentage of scalar instructions, which could be otherwise significant. It is also worth noting that, in some problems, invariant sub-expressions along $j$ are identical to those along $k$ (e.g. in Helmholtz). In these cases, we safely avoid redundant pre-computation since, as anticipated in Section~\ref{sec:background}, a property of our domain is that $j$ and $k$ loops share the same iteration space.

@@

% Register allocation and register locality
\subsection{Model-driven Vector-register Tiling}
One notable problem of assembly kernels concerns register allocation and register locality. The critical situation occurs when loop trip counts and accessed variables are such that the vector-registers pressure is high. Since the kernel's working set fits the L1 cache, it is remarkably important to optimize register management. Canonical optimizations, such as loop interchange, unroll, and unroll-and-jam, can be employed to deal with this problem. In COFFEE, these optimizations are supported either by means of explicit code transformations (interchange, unroll-and-jam) or indirectly by delegation to the compiler through standard pragmas (unroll). Tiling at the level of vector registers is an additional feature of COFFEE. Based on the observation that the evaluation of the element matrix can be reduced to a ``summation of outer products'' along the $j$ and $k$ dimensions, a model-driven vector-register tiling strategy can be implemented. If we consider the code snippet in Figure~\ref{code:helmholtz-licm} and we ignore the presence of the operation \texttt{det*W3[i]}, the computation of the element matrix is abstractly expressible as
\begin{equation}
\label{outer-product}
A_{jk} = \sum_{\substack{
  x \in B' \subseteq B \\
  y \in B'' \subseteq B}}
x_j\cdot y_k ~~~~~~ j,k = 0,...,3
\end{equation}
where $B$ is the set of all basis functions (or temporary variables, e.g. \texttt{LI$\_$0}) accessed in the kernel, whereas $B'$ and $B''$ are generic problem-dependent subsets. Regardless of the specific input problem, and by abstracting from the presence of all variables independent of both $j$ and $k$, the element matrix computation is always reducible to this form. Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 entries ($j,k=0,...,4$) of the element matrix using just 2 vector registers, which represent a 4$\times$4 tile, assuming $\vert B' \vert = \vert B'' \vert = 1$. Values in a register are shuffled each time a product is performed. Standard compiler auto-vectorization ($gnu$ and $intel$), instead, executes 4 broadcast operations (i.e. ``splat'' of a value over all of the register locations) along the outer dimension to perform the calculation. Besides a larger number of cache accesses, it needs to keep between $f=1$ and $f=3$ extra registers to perform the same 16 evaluations when unroll-and-jam is used, with $f$ being the unroll-and-jam factor.

@@

\begin{figure}[t]
\centerline{\includegraphics[scale=0.7]{Pictures/vect-restore.pdf}}
\caption{Restoring the storage layout after \emph{op-vect}. The figure shows how 4$\times$4 elements in the top-left block of the element matrix $A$ can be moved to their correct positions. Each rotation, represented by a group of three same-colored arrows, is implemented by a single shuffle intrinsics.}
\label{fig:restore-layout}
\end{figure}

@@

% Splitting
\subsection{Expression Splitting}
\label{sec:expr-split}
In complex kernels, like Burgers in Figure~\ref{code:burgers}, and on certain architectures, achieving effective register allocation can be challenging. If the number of variables independent of the innermost-loop dimension is close to or greater than the number of available CPU registers, poor register reuse is likely to be obtained. This usually happens when the number of basis function arrays, temporaries introduced by generalized loop-invariant code motion, and problem constants is large. For example, applying loop-invariant code motion to Burgers on a 3D mesh needs 33 temporaries for the \texttt{ijk} loop order, so compiler's hoisting of invariant loads out of the $k$ loop can be inefficient on architectures with a relatively low number of registers. One potential solution to this problem consists of suitably ``splitting'' the computation of the element matrix $A$ into multiple sub-expressions; an example, for the Helmholtz problem, is given in Figure~\ref{code:helmholtz-split}.

Splitting an expression has three drawbacks. Firstly, it increases the number of accesses to $A$ proportionally to the ``split factor'', which is the number of sub-expressions produced. Secondly, depending on how the split is executed, it can lead to redundant computation (e.g. the product $det*W3[i]$ is performed times number of sub-expressions in the code of Figure~\ref{code:helmholtz-split}). Finally, it might affect register locality: for instance, the same array could be accessed in different sub-expressions, requiring a proportional number of loads be performed. This is not the case of the Helmholtz example. Nevertheless, as shown in Section~\ref{sec:perf-results}, the performance gain from improved register reuse along inner dimensions can still be greater, especially if the split factor and the splitting itself use heuristics to minimize the aforementioned issues.

@@

Firedrake provides users with the Unified Form Language to write problems in a notation resembling mathematical equations. At run-time, the high-level specification is translated by the FEniCS Form Compiler~\cite{FFC-Compiler} into an abstract syntax tree (AST) representation of one or more finite element assembly kernels. ASTs are then passed to COFFEE, which is capable of applying the transformations described in Section~\ref{sec:code-transf}. The output of COFFEE, C code, is eventually provided to PyOP2~\cite{pyop2isc}, where just-in-time compilation and execution over the discretized domain take place. Because of the large number of parametric transformations, COFFEE needs a mechanism to select the most suitable optimization strategy for a given problem. Auto-tuning might be used, although it would significantly increase the run-time overhead, since the generation of ASTs occurs at run-time as soon as problem-specific data are available. Our optimization strategy, based on heuristics and a simple cost model, is described in the following, along with an overview of the compiler.

@@

All loops are interchangeable provided that temporaries are introduced if the nest is not perfect. For the employed storage layout, the loop permutations \texttt{ijk} and \texttt{ikj} are likely to maximize performance. Conceptually, this is motivated by the fact that if the $i$ loop were in an inner position, then a significantly higher number of load instructions would be required at every iteration. We tested this hyphotesis in manually crafted kernels. We found that the performance loss is greater than the gain due to the possibility of accumulating increments in a register, rather than in memory, along the $i$ loop. The choice between \texttt{ijk} and \texttt{ikj} depends on the number of load instructions that can be hoisted out of the innermost dimension. Our compiler chooses the loop along which the number of invariant loads is smaller as outer so that more registers remain available to carry out the computation of the element matrix.

Loop unroll and unroll-and-jam of outer loops are fundamental to expose instruction-level parallelism, so tuning critical parameters, like the unroll factor, is of great importance. However, from our experience of inspecting assembly code and comparing with other hand-written implementations, recent versions of a vendor compiler like Intel's employ cost models capable of estimating close-to-optimal values for such parameters. This is particularly true for assembly kernels, where the loop nest is affine, bounds are usually very small and known at compile-time, and memory accesses are stride-1. We therefore leave the backend compiler in charge of selecting the unroll factor. This choice also simplifies COFFEE's cost model. The only situation in which we explicitly unroll-and-jam a loop is when \emph{op-vect} is used, since the transformed code seems to prevent the Intel compiler from applying this optimization, even if specific pragmas are added.

%Note that, regardless of the output of the cost model, the unroll-and-jam factor never exceeds the actual size of the outer loop (line 19 in Figure~\ref{algo:PyOP2Compiler}, assuming the default loop order \texttt{ijk}).
